Key Drivers of Community Garden Growth: Analyzing Trends in PollinateTO Funding

### 2024-12-03 paper/paper.qmd
Update Paper

---
title: "Key Drivers of Community Garden Growth: Analyzing Trends in PollinateTO Funding Across Toronto’s Neighborhoods"
subtitle: "A Bayesian Approach to Understanding the Impact of Time, Type, and Location on Garden Size and Number."
author: 
  - Aliza Abbas Mithwani
thanks: "Code and data are available at: [https://github.com/alizamithwani/PollinateTO.git]"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

#### Workspace setup ####
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(arrow)
```

@@

# Data {#sec-data}
## Software Packages
This analysis utilizes the R programming language [@citeR], with the 'rstanarm' [@rstanarm] and 'tidyverse' [@tidyverse] packages for statistical modeling and data manipulation. Our data comes from the PollinateTO Primary Project Garden Locations dataset [@PTdata], which is hosted on Toronto's Open Data Portal. Following the approach outlined in @tellingstories, we leverage these tools to explore patterns and trends in the allocation of PollinateTO funding.

@@

## Methodology
The PollinateTO dataset collects detailed information on community-led garden projects funded by Toronto’s PollinateTO initiative. Data is sourced from the city’s Open Data portal and includes variables like garden type, location, funding year, and size, enabling analysis of trends in urban biodiversity projects. The dataset uses standardized attributes and is updated annually, making it a valuable tool for evaluating the initiative’s impact across various Toronto neighborhoods, though it may be subject to self-reporting biases and geographical limitations in precision.

## Data Visualization
To understand the data provided better, we visualize it using tables and graphs. To begin with, let's see how many gardens each type of neighborhood (NIA, EN or None) has and separate them by whether or not they are indigenous-led in @fig-IndNeiBoxPlot and @fig-IndNeiBarChart.

```{r}
#| label: fig-IndNeiBoxPlot
#| fig-cap: Summary of Gardens by Neighborhood Type
#| echo: false

# Summarize the data
PT_summary_table <- PT_analysis_data %>%
  group_by(nia_or_en) %>%
  summarise(
    Total_Gardens = n(), # Total number of gardens in each neighborhood type
    Indigenous_Gardens = sum(is_indigenous_garden), # Count of Indigenous-led gardens
    Non_Indigenous_Gardens = Total_Gardens - Indigenous_Gardens, # Count of Non-Indigenous gardens
    Proportion_Indigenous = (Indigenous_Gardens / Total_Gardens) * 100, # Proportion of Indigenous gardens
    .groups = "drop"
  )

@@

```{r}
#| label: fig-IndNeiBarChart
#| fig-cap: Distribution of Garden Types Funded Over Time
#| echo: false
# Filter for Indigenous-Led gardens only
PT_indigenous <- PT_analysis_data %>%
  filter(is_indigenous_garden == 1) %>%
  group_by(nia_or_en) %>%
  summarise(count = n(), .groups = "drop")

# Create a bar chart
ggplot(PT_indigenous, aes(x = nia_or_en, y = count, fill = nia_or_en)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Count of Indigenous-Led Gardens by Neighborhood Type",
    x = "Neighborhood Type (NIA or EN)",
    y = "Number of Indigenous-Led Gardens",
    fill = "Neighborhood Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

Next, we visualize the trend of garden sizes and numbers across time to examine the impact of the initiative over time in @fig-trendline.

@@

# Create the line chart with two y-axes (total gardens and average size)
ggplot(PT_summary, aes(x = year_funded)) +
  geom_line(aes(y = total_gardens), color = "blue", linewidth = 1.2, linetype = "solid") + 
  geom_line(aes(y = avg_garden_size), color = "red", linewidth = 1.2, linetype = "dashed") +
  scale_y_continuous(
    name = "Total Number of Gardens", 
    sec.axis = sec_axis(~., name = "Average Garden Size (Estimated)")
  ) +
  labs(
    x = "Year Funded",
    y = "Total Number of Gardens"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "red"),
    axis.text.y.right = element_text(color = "red")
  )
```
\newpage
Now, let's analyze the proportion of each garden type visually across time in @fig-distributiontype to see if certain garden types increased in number and proportion overtime, reflecting shifting funding priorities and initiatives.

@@

# Create the bar chart
ggplot(PT_summary, aes(x = year_funded, y = garden_count, fill = garden_type)) +
  geom_bar(stat = "identity", position = "stack") + # Stacked bar chart
  scale_fill_brewer(palette = "Set3") + # Use a color palette for clarity
  labs(
    x = "Year Funded",
    y = "Number of Gardens",
    fill = "Garden Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 1) # Rotate x-axis labels for readability
  )
```
\newpage
To prevent outliars from making graphs unreadable, we use log-transformed garden sizes to analyze differences in garden sizes based on neighborhood type in @fig-neighborhoodsize. A boxplot is used to do this where the 3 neighborhood categories are EN (Emerging Neighborhoods), NIA (Neighborhood Improvement Areas) and None (neither EN nor NIA).

```{r}
#| label: fig-neighborhoodsize
#| fig-cap: Log-Transformed Garden Sizes by Neighborhood Type
#| echo: false
# Add a log-transformed column for garden size
PT_analysis_data <- PT_analysis_data %>%
  mutate(log_garden_size = log1p(estimated_garden_size)) # log1p handles zero values safely

# Use the log-transformed column in the boxplot
ggplot(PT_analysis_data, aes(x = nia_or_en, y = log_garden_size, fill = nia_or_en)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, notch = FALSE) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    x = "Neighborhood Type (NIA or EN)",
    y = "Log of Estimated Garden Size",
    fill = "Neighborhood Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```
# Model
In the analysis, I used a Bayesian Logistic Regression Model to examine the relationship between garden size and other factors like the type of garden, the year the project was funded, the neighborhood type (emerging neighborhood, EN vs neighborhood improvement area, NIA), and if it is an indigenous-led project. We are interested in modeling whether a garden is large (binary outcome) as a function of various factors such as the garden type, year funded, neighborhood type, indigenous garden status, and ward name. Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

```{Latex}
\begin{align}
y_i|\pi_i &\sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) &= \alpha + \beta_1 \times \mbox{garden_type}_i + \beta_2 \times \mbox{year_funded}_i + \beta_3 \times \mbox{nia_or_en}_i + \beta_4 \times \mbox{is_indigenous_garden}_i + \beta_5 \times \mbox{ward_name}_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
\beta_4 &\sim \mbox{Normal}(0, 2.5) \\
\beta_5 &\sim \mbox{Normal}(0, 2.5)
\end{align}
```

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

@@

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

#modelsummary::modelsummary(
#  list(
#    "First model" = first_model
 # ),
 # statistic = "mad",
 # fmt = 2
#)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

@@

@@

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

#pp_check(first_model) +
#  theme_classic() +
#  theme(legend.position = "bottom")

#posterior_vs_prior(first_model) +
#  theme_minimal() +
 # scale_color_brewer(palette = "Set1") +
#  theme(legend.position = "bottom") +
#  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

#plot(first_model, "trace")

#plot(first_model, "rhat")
```

>>>>

---
title: "Key Drivers of Community Garden Growth: Analyzing Trends in PollinateTO Funding"
subtitle: "A Bayesian Approach to Understanding the Impact of Time, Type, and Location on Garden Size and Number in Toronto's Neighborhoods from 2020 to 2023."
author: 
  - Aliza Abbas Mithwani
thanks: "Code and data are available at: [https://github.com/alizamithwani/PollinateTO.git]"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format:
  pdf:
    toc: true
number-sections: true
header-includes:
  - \usepackage{pdflscape}
bibliography: references.bib
---

@@

#### Workspace setup ####
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(arrow)
library(rstanarm)
library(patchwork)
library(tinytable)
library(kableExtra)
library(broom.mixed)

#### Read Model ####
large_garden_model <- readRDS("../models/large_garden_model.rds")
```

@@

# Data {#sec-data}
## Software Packages
This analysis utilizes the R programming language [@citeR], with the `rstanarm` [@rstanarm] and `tidyverse` [@tidyverse] packages for statistical modeling and data manipulation. Our data comes from the PollinateTO Primary Project Garden Locations dataset [@PTdata], which is hosted on Toronto's Open Data Portal. Following the approach outlined in @tellingstories, we leverage these tools to explore patterns and trends in the allocation of PollinateTO funding.

@@

## Methodology
The PollinateTO dataset collects detailed information on community-led garden projects funded by Toronto’s PollinateTO initiative. Data is sourced from the city’s Open Data portal and includes variables like garden type, location, funding year, and size, enabling analysis of trends in urban biodiversity projects. The dataset uses standardized attributes and is updated annually, making it a valuable tool for evaluating the initiative’s impact across various Toronto neighborhoods, though it may be subject to self-reporting biases and geographical limitations in precision.
\newpage

## Data Visualization
To understand the data provided better, we visualize it using tables and graphs. To begin with, let's see the distribution of Indigenous gardens in each type of neighborhood (NIA, EN or None) in @tbl-IndNei and @fig-IndNeiBarChart. As seen in @tbl-IndNei, Indigenous gardens take up a higher proportion of all gardens located in Emerging Neighborhoods; however, there are significantly less Emerging Neighborhood gardens than Neighborhood Improvement Area gardens. @fig-IndNeiBarChart helps us realize that overall, there aren't a lot of Indigenous(-led) gardens in this initiative. There is only 1 in an Emerging Neighborhood, 1 in a Neighborhood Improvement Area and 3 in other neighborhoods (not EN or NIA).

```{r}
#| label: tbl-IndNei
#| tbl-cap: Proportion of Indigenous Gardens in Each Neighborhood Type
#| echo: false

# Summarize the data
PT_summary_table <- PT_analysis_data %>%
  group_by(nia_or_en) %>%
  summarise(
    Total_Gardens = n(), # Total number of gardens in each neighborhood type
    Indigenous_Gardens = sum(is_indigenous_garden), # Count of Indigenous-led gardens
    Non_Indigenous_Gardens = Total_Gardens - Indigenous_Gardens, # Count of Non-Indigenous gardens
    Proportion_Indigenous = round((Indigenous_Gardens / Total_Gardens) * 100,2), # Proportion of Indigenous gardens
    .groups = "drop"
  )

@@

```{r}
#| label: fig-IndNeiBarChart
#| fig-cap: Number of Indigenous Gardens Funded Across Neighborhoods
#| echo: false
# Filter for Indigenous-Led gardens only
PT_indigenous <- PT_analysis_data %>%
  filter(is_indigenous_garden == 1) %>%
  group_by(nia_or_en) %>%
  summarise(count = n(), .groups = "drop")

# Create a bar chart
ggplot(PT_indigenous, aes(x = nia_or_en, y = count, fill = nia_or_en)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Count of Indigenous-Led Gardens by Neighborhood Type",
    x = "Neighborhood Type (NIA or EN)",
    y = "Number of Indigenous-Led Gardens",
    fill = "Neighborhood Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```
\newpage
Next, we visualize the trend of garden sizes and numbers across time to examine the impact of the initiative over time in @fig-trendline. The total number of new gardens funded by PollinateTO decreased significantly from 2020 to 2021, but then increased over 2021 to 2023 by a number greater than the initial drop. Estimated Average Garden Size has been decreasing significantly from 2020 to 2022, but recently increase by a small amount from 2022 to 2023.

@@

# Create the line chart with two y-axes (total gardens and average size)
ggplot(PT_summary, aes(x = year_funded)) +
  geom_line(aes(y = total_gardens), color = "blue", linewidth = 1.2, linetype = "solid") + 
  geom_line(aes(y = avg_garden_size), color = "red", linewidth = 1.2, linetype = "dashed") +
  scale_y_continuous(
    name = "Total Number of Gardens", 
    sec.axis = sec_axis(~., name = "Average Garden Size (Estimated)")
  ) +
  labs(
    x = "Year Funded",
    y = "Total Number of Gardens"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "red"),
    axis.text.y.right = element_text(color = "red")
  )
```
\newpage
Now, let's analyze the proportion of each garden type across time in @fig-distributiontype to see if certain garden types increased in proportion overtime, reflecting shifting funding priorities and initiatives. School Learning Gardens were consistently the most abundant type of garden compared to the rest in each year, and grew in proportion consistently from 2020 to 2023. City Park Gardens used to be the second most abundant type of garden in 2020, 2021, and 2022, but they recently fell significantly in proportion and were overtaken by Multiple Front Yard Gardens.

@@

# Create the bar chart
ggplot(PT_summary, aes(x = year_funded, y = garden_count, fill = garden_type)) +
  geom_bar(stat = "identity", position = "stack") + # Stacked bar chart
  scale_fill_brewer(palette = "Set3") + # Use a color palette for clarity
  labs(
    x = "Year Funded",
    y = "Number of Gardens",
    fill = "Garden Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 1) # Rotate x-axis labels for readability
  )
```
\newpage
To prevent outliars from making graphs unreadable, we use log-transformed garden sizes to analyze differences in garden sizes based on neighborhood type in @fig-neighborhoodsize. A dot plot is used to do this where the 3 neighborhood categories are EN (Emerging Neighborhoods), NIA (Neighborhood Improvement Areas) and None (neither EN nor NIA).

```{r}
#| label: fig-neighborhoodsize
#| fig-cap: Log-Transformed Garden Sizes by Neighborhood Type
#| echo: false
#| warning: false

# Add a log-transformed column for garden size
PT_analysis_data <- PT_analysis_data %>%
  mutate(log_garden_size = log1p(estimated_garden_size)) # log1p handles zero values safely

# Use the log-transformed column in the dotplot
ggplot(PT_analysis_data, aes(x = nia_or_en, y = log_garden_size, fill = nia_or_en)) +
  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.8) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Log-Transformed Garden Sizes by Neighborhood Type",
    x = "Neighborhood Type (NIA or EN)",
    y = "Log of Estimated Garden Size",
    fill = "Neighborhood Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
# Model
In the analysis, I used a Bayesian Logistic Regression Model to examine the relationship between garden size and other factors like the type of garden, the year the project was funded, the neighborhood type (emerging neighborhood, EN vs neighborhood improvement area, NIA), and if it is an indigenous-led project. We are interested in modeling whether a garden is large (binary outcome) as a function of various factors such as the garden type, year funded, neighborhood type, indigenous garden status, and ward name. Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

\begin{align} 
y_i|\pi_i & \sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) &= \alpha 
    + \beta_1 \times \mbox{garden type}_i 
    + \beta_2 \times \mbox{year funded}_i \notag \\
    &\quad + \beta_3 \times \mbox{nia or en}_i 
    + \beta_4 \times \mbox{is indigenous garden}_i 
    + \beta_5 \times \mbox{ward name}_i \notag \\
    &\quad + \beta_6 \times (\mbox{garden type}_i \times \mbox{nia or en}_i) \\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
\beta_4 &\sim \mbox{Normal}(0, 2.5) \\
\beta_5 &\sim \mbox{Normal}(0, 2.5)
\end{align}

This Bayesian Logistic Regression model predicts whether a garden is classified as "large" ($y_i$), a binary outcome modeled as $y_i \sim \mbox{Bern}(\pi_i)$, where $\pi_i$ is the probability of the garden being large. The predictors include: garden type, a categorical variable indicating the type of garden (e.g., pollinator, vegetable); year funded, a continuous variable representing the year the garden received funding; nia or en, a categorical variable indicating whether the garden is in a Neighborhood Improvement Area (NIA) or elsewhere (EN); is indigenous garden, a binary variable identifying whether the garden is indigenous-led; and ward name, a categorical variable representing the geographic ward. Additionally, this model includes an interaction term between garden type and nia or en, which helps examine whether the effect of garden type on the likelihood of being large differs depending on whether the garden is in an NIA or EN neighborhood. These predictors influence $\pi_i$ through the logit link function: $\mbox{logit}(\pi_i) = \alpha + \beta_1 \times \mbox{garden type}_i + \beta_2 \times \mbox{year funded}_i + \beta_3 \times \mbox{nia or en}_i + \beta_4 \times \mbox{is indigenous garden}_i + \beta_5 \times \mbox{ward name}_i + \beta_6 \times (\mbox{garden type}_i \times \mbox{nia or en}_i)$, where $\alpha$ is the intercept, $\beta_k$ are the coefficients, and $\beta_6$ represents the interaction term between garden type and nia or en. Priors for $\alpha$ and $\beta_k$ are defined as Normal(0,2.5), reflecting weakly informative assumptions. We run the model in R [@citeR] using the `rstanarm` package of @rstanarm.

@@

The Bayesian logistic regression model employed in this analysis is appropriate for investigating the factors influencing the likelihood of a garden being classified as large based on binary outcomes. Specifically, this model explores how various predictors—such as garden type, year funded, whether the garden is indigenous-led, the neighborhood classification (NIA or EN), and ward names—affect garden size. Given the binary nature of the outcome variable (is_large_garden), logistic regression is the suitable choice as it models probabilities between 0 and 1. The inclusion of an interaction term between garden type and nia or en further enriches the analysis, allowing us to examine whether the effect of garden type on garden size differs by neighborhood type. This is especially relevant as different neighborhoods may have different conditions and priorities that could influence the type and size of gardens.

The Bayesian framework provides several advantages, including the ability to incorporate prior beliefs about the parameters, produce credible intervals, and assess model uncertainty. Furthermore, Bayesian methods enable a more robust assessment of convergence through posterior predictive checks and the examination of diagnostic tools such as trace and Rhat plots, ensuring reliable inferences.

The choice of the Bayesian approach in this context also allows for the flexible inclusion of multiple categorical predictors, which is crucial for this study, as it aims to assess the relationship between categorical factors like garden type, ward names, and the interaction between garden type and nia or en with a quantitative outcome. The model's use of informative priors helps regularize the estimates, particularly when dealing with categorical predictors that may have complex relationships with the outcome variable. Additionally, the Bayesian methodology provides the opportunity to assess model diagnostics comprehensively, ensuring that the model fits well with the data and that the parameters are estimated with high precision. Overall, this model is well-suited to capture the complexities of the data while offering interpretability and flexibility through its probabilistic nature.

@@

Our results are summarized in @tbl-modeltable. The Bayesian logistic regression model reveals several important relationships between garden characteristics and the likelihood of a garden being classified as large. Among garden types, "Multiple Property Garden" stands out with a substantial positive coefficient (14.94), suggesting that such gardens are strongly associated with larger sizes. Conversely, "Boulevard Garden, Rain Garden" exhibits a notable negative coefficient (-13.77), indicating a reduced likelihood of being large. Other garden types, such as "Communal Garden" and "School Learning Garden," have smaller coefficients, implying a weaker or negligible relationship with garden size. This variability in the influence of garden type highlights the importance of understanding project-specific features when predicting garden outcomes.

The neighborhood context also plays a significant role. Gardens located in areas classified as "None" (neither EN nor NIA) have a relatively high positive coefficient (4.54), suggesting that these gardens are more likely to be larger compared to those in ENs or NIAs. Additionally, Indigenous-led gardens are associated with a negative coefficient (-8.86), suggesting they may be less likely to be large. The interaction terms, such as those between "Multi-Residential Garden" and neighborhood type, indicate that specific combinations of garden type and neighborhood characteristics can further modify the likelihood of a garden being large. These findings underscore the complexity of garden size determinants and suggest targeted strategies may be needed to ensure equitable garden funding and development.

```{r}
#| echo: false
#| eval: true
#| label: tbl-modeltable
#| warning: false
#| tbl-cap: "Explanatory model of whether a garden is large based on certain factors: Indigenous-led indicator, Year Funded, Neighborhood, Garden type and Ward Names"

# Create the table data
model_table <- modelsummary::modelsummary(
  list("Model" = large_garden_model),
  statistic = "mad",
  fmt = 2,
  output = "data.frame" # Output as a data frame for kable compatibility
)

# Render the table using kableExtra
model_table %>%
  kbl() %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>%
  landscape()
```
@fig-model-coefficients visualizes the estimated effects of various predictors on the likelihood of a garden being classified as "large" in the PollinateTO initiative. Each point represents the coefficient estimate for a predictor, indicating the direction and strength of its relationship with garden size. Predictors with positive coefficients increase the probability of a garden being large, while those with negative coefficients decrease it. The horizontal bars show the 95% credible intervals, capturing the uncertainty in the estimates. Predictors whose credible intervals do not cross the dashed red line at zero are likely to have significant effects. This plot highlights which variables, such as garden type, neighborhood classification, or year funded, are most influential in determining garden size, providing valuable insights into the factors driving outcomes in the program.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-model-coefficients
#| fig-cap: Coefficients of the Model
#| fig-height: 8

# Coefficients for the model
coefficients <- broom.mixed::tidy(large_garden_model, conf.int = TRUE)

# Plot for the coefficients of the model
ggplot(coefficients, aes(x = estimate, y = term)) +
  geom_point() + # Coefficient points
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) + # Confidence/Credible interval bars
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + # The zero line
  labs(x = "Coefficient Estimate", y = "Predictor Variable") +
  theme_minimal()
``` 
\newpage
## Predictive Probability
The bar chart in @fig-predictiveprobability visualizes the predicted probability that a garden funded by PollinateTO will be classified as "large," based on the combination of garden type (garden_type) and neighborhood classification (nia_or_en). The height of each bar represents the average predicted probability for each garden type and neighborhood combination, with all other variables (such as year funded, Indigenous-led status, and ward) held constant. The chart helps identify which garden types and neighborhood categories are most likely to result in larger gardens. By examining the differences in predicted probabilities across the garden types and neighborhood types, we can discern patterns in how these factors influence garden size. This analysis can inform decisions about resource allocation, funding strategies, or identifying areas where specific types of gardens are more likely to grow larger, contributing to the broader goals of the PollinateTO program.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-predictiveprobability
#| fig-cap: Predictive Probability of Large Gardens
#| fig-height: 8
#| fig-width: 16
#### Predictive Probability Visualization ####

# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(rstanarm)

# Re-factor variables in the dataset
PT_analysis_data <- PT_analysis_data %>%
  mutate(
    garden_type = as.factor(garden_type),
    nia_or_en = as.factor(nia_or_en),
    is_indigenous_garden = as.factor(is_indigenous_garden),
    ward_name = as.factor(ward_name)
  )

# Generate new data for prediction
new_data <- expand.grid(
  garden_type = levels(PT_analysis_data$garden_type),
  nia_or_en = levels(PT_analysis_data$nia_or_en),
  year_funded = median(PT_analysis_data$year_funded, na.rm = TRUE),
  is_indigenous_garden = levels(PT_analysis_data$is_indigenous_garden)[1], # Hold constant
  ward_name = levels(PT_analysis_data$ward_name)[1] # Hold constant
)

# Check if new_data is non-empty
if (nrow(new_data) == 0) stop("new_data is empty. Check factor levels and variables used in expand.grid.")

# Predictive probabilities from the posterior
predicted_probs <- rstanarm::posterior_predict(large_garden_model, newdata = new_data, type = "response")

# Summarize posterior predictions for each row in new_data
summary_probs <- as.data.frame(t(predicted_probs)) %>% # Transpose to match new_data rows
  rowwise() %>%
  summarise(
    mean_prob = mean(c_across(everything())),
    lower_prob = quantile(c_across(everything()), probs = 0.025),
    upper_prob = quantile(c_across(everything()), probs = 0.975)
  )

# Combine summary with new_data
new_data <- bind_cols(new_data, summary_probs)

# Plot the predicted probabilities
ggplot(new_data, aes(x = garden_type, y = mean_prob, fill = nia_or_en)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  labs(
    x = "Garden Type",
    y = "Predicted Probability",
    fill = "Neighborhood Type (NIA or EN)"
  ) +
  theme_minimal() +
  theme(
    axis.title.y = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12), # Rotates labels by 45 degrees
    axis.text.y = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)   
  )
```

\newpage
# Discussion

## Unearthing Patterns in Garden Allocation {#sec-first-point}
This paper investigates the factors influencing the size of gardens funded under Toronto’s PollinateTO initiative using a Bayesian logistic regression model. Specifically, it explores how garden size (a binary outcome of “large” vs. “not large”) is associated with variables such as garden type, year funded, neighborhood classification (Emerging Neighborhoods or Neighborhood Improvement Areas), Indigenous-led status, and ward-level geographic differences. By leveraging this model, the analysis identifies patterns and disparities in garden allocation, providing insights into which factors are most predictive of larger gardens. The approach incorporates interaction effects to capture nuanced relationships between garden types and neighborhood characteristics.

The inclusion of credible intervals from the Bayesian model ensures a robust interpretation of uncertainty in coefficient estimates. Additionally, key goodness-of-fit measures (e.g., R², WAIC) are evaluated, demonstrating the model’s ability to explain nearly half of the variation in garden size. These findings offer a data-driven lens through which urban sustainability programs like PollinateTO can be assessed and improved.

## Neighborhood Inequities in Urban Gardening

One of the central insights from this analysis is the existence of disparities in garden size across neighborhoods. Specifically, gardens in areas classified as “None” (neither EN nor NIA) are significantly more likely to be large, while gardens in NIAs face challenges in achieving similar sizes. This suggests that systemic factors, such as socioeconomic inequalities, land availability, and access to resources, influence the outcomes of urban greening initiatives.

This insight highlights a broader inequity in urban sustainability programs: neighborhoods with fewer resources or systemic disadvantages may struggle to achieve comparable environmental benefits. This finding emphasizes the importance of embedding equity-focused considerations into grant allocation processes. Addressing these disparities can improve not only the environmental impact of such initiatives but also their social relevance and inclusivity.

## Redefining Success for Indigenous-Led Projects

Indigenous-led projects are less likely to result in large gardens, pointing to potential barriers that Indigenous communities face in participating fully in urban greening initiatives. This finding underscores the need to reevaluate the criteria for “success” in such programs. Indigenous-led projects may prioritize cultural and ecological goals—such as the preservation of native plant species or community education—over garden size.

This raises an important consideration for future urban sustainability programs: qualitative impacts, such as community engagement or cultural significance, may be equally (if not more) important than quantitative measures like garden size. By failing to account for these dimensions, existing frameworks risk marginalizing projects that offer critical social and cultural benefits. This insight encourages a shift in how success is defined and measured in urban greening efforts.

# Weaknesses and next steps
I will address some of the weaknesses and limitations of the current analysis, focusing on areas where the methodology or data could be improved. Following this, I will explore potential next steps, outlining opportunities for future research and policy development that build on these findings to create more equitable and effective urban sustainability initiatives. This section aims to critically reflect on the study while charting a path forward for both academic inquiry and practical applications.

## Challenges and Limitations in Modeling Garden Outcomes
While the model provides valuable insights, it is not without limitations. First, the binary categorization of garden size as “large” vs. “not large” may oversimplify the complexity of garden outcomes. A more nuanced approach, such as modeling size as a continuous variable or including qualitative measures of success, might provide richer insights.

Second, the dataset is limited to gardens funded by PollinateTO, which may not fully represent the broader urban gardening landscape in Toronto. Selection bias could influence the generalizability of these findings to other urban greening initiatives. Additionally, the potential for omitted variable bias exists, as factors like community engagement levels or prior gardening experience were not included in the analysis but may significantly impact garden outcomes.

Finally, while interaction terms capture some complexity, they add interpretive challenges, as not all combinations of garden types and neighborhoods had sufficient data to provide precise estimates. Future work could address this by gathering larger, more diverse datasets or employing hierarchical models to account for varying group sizes.

## Opportunities for Future Research and Policy Improvements
Several open questions remain. One key area for future research is the inclusion of community-level variables, such as demographic characteristics, access to volunteer networks, and proximity to green spaces. These factors may help explain why certain neighborhoods are more likely to host large gardens.

Additionally, further work is needed to explore the qualitative outcomes of urban gardens, particularly for Indigenous-led projects. Incorporating interviews, case studies, or community feedback could provide a deeper understanding of how these gardens contribute to cultural preservation and community well-being.

From a methodological perspective, future studies could explore alternative modeling techniques, such as hierarchical Bayesian models, to better account for ward-level variability and cross-neighborhood interactions. Expanding the dataset to include other urban greening initiatives or longitudinal data could also provide insights into how garden outcomes evolve over time.

Lastly, there is a need to test the effectiveness of policy interventions informed by these findings. For instance, targeted funding for NIAs or revised evaluation criteria for Indigenous-led projects could be piloted and evaluated to determine their impact on garden size and community engagement. This iterative process of evidence-based policy design and evaluation can help create more equitable and effective urban sustainability programs.

@@

## Data Cleaning

## Data Simulation

@@

In @fig-ppcheck we implement a posterior predictive check. The posterior predictive check (pp_check) helps assess how well the model fits the observed data in the context of your research on the factors influencing community garden funding by PollinateTO across different Toronto neighborhoods. By comparing the model’s predictions (posterior distribution) with the observed data, this diagnostic reveals whether the model is accurately capturing the key influences, such as garden type, funding year, and neighborhood characteristics. A good fit suggests that these factors are well-represented in the model, while significant deviations between the predictions and the data may indicate that the model needs refinement. This check is essential for validating the model’s effectiveness in answering your research question.

```{r}
#| label: fig-ppcheck
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(large_garden_model) +
  theme_classic() +
  theme(legend.position = "bottom")
```
\newpage
In @fig-posteriorvsprior we compare the posterior with the prior. The posterior vs. prior comparison visualizes how the model’s posterior estimates for various parameters compare to the prior distributions. In the context of your research on the factors influencing the size and number of community gardens funded by PollinateTO, this comparison helps evaluate whether the data has significantly updated the prior beliefs embedded in the model. A well-calibrated model should show a substantial difference between the prior and posterior for key parameters, indicating that the observed data has influenced the model’s estimates. This comparison is crucial for understanding which variables are most influential in determining garden size and funding across neighborhoods.

Given the large number of parameters in the model, a random sample of 5 garden types and 5 ward names was taken for this visualization. This reduces complexity and ensures the comparison remains interpretable, focusing on a manageable subset of the factors that may affect the size and number of community gardens across Toronto.

```{r}
#| label: fig-posteriorvsprior
#| fig-cap: Compare the posterior with the prior 
#| eval: true
#| echo: false
#| warning: false
#| message: false

PT_analysis_data <- PT_analysis_data %>%
mutate(is_large_garden = ifelse(estimated_garden_size > mean(estimated_garden_size, na.rm = TRUE), 1, 0))

garden_type_list <- c("garden_typeBoulevard Garden, Rain Garden","garden_typeCity Park","garden_typeCity Park, Rain Garden","garden_typeCommunal Garden","garden_typeMulti-Residential Garden","garden_typeMultiple Boulevard Gardens","garden_typeMultiple Front Yard Gardens","garden_typeMultiple Property Garden","garden_typeSchool Learning Garden","garden_typeSpiritual Centre Garden","garden_typeSpiritual Centre Garden, Rain Garden")
set.seed(522)
garden_type_sample <- sample(garden_type_list,5)
